What is an Ancestral Recombination Graph?

### 2022-02-01 paper.tex
Section on phylogenetic networks, edits to section on parsimonious ARG reconstruction (#48)
* Added section on phylogenetic networks; edited section on parsimonious ARG reconstruction

* Added some comments

* Adding back .svg figs
Working through background material
Fix wu2011 citation

\section*{The ARG as a data structure}

@@

\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.

@@

% HYW: this is a rough first pass. We should take care not to assume that the tree sequence
% approach is the only encoding

The other use of the the term ARG, and the one for which we suggest the term be
reserved, is to describe the *structure* resulting from the genetic process of
inheritance. This structure, like any other graph, consists of *nodes*
connected by *edges*. In the context of genetics, the nodes represent (haploid)
genomes, and the edges represent paths of inheritance. Although for theoretical
purposes a graph like this can be created by backwards-in-time simulations
using the CwR, in reality it is created forwards-in-time, as a population
evolves.

Since genetic inheritance is unidirectional, the edges in this graph structure
are *directed*, with a "parent" and "child" node. Moreover, there is a strict
temporal order, so that children cannot be their own parents: in other words,
there are no cycles in the graph, and it is technically a form of "directed
acyclic graph" (DAG).

% HYW: NB: worded carefully below, as it should also be possible to encode ARGs even if there is not a single fixed coordinate system

There is one key feature which distinguishes an ARG for other graphs. Although
a genome (graph node) may have more than one parent, the particulate nature of
genetic inheritance means that each letter in the genome can only come from a
*single* parent. Hence if we compare the equivalent letters from multiple
extant genomes, their paths of inheritance must form a *tree* (which,
incidentally, is why phylogenetics plays such an important role in evolutionary
theory). As we shall argue, the ability to generate these "local trees" from an
ARG is of key importance.

When describing an ARG, the primary issue, from a practical point of view, is
how to annotate the graph with the details of which piece of genome has come
from which parent. There are two possibilities: the annotations can be
associated either with the nodes, or with the edges.

@@

\subsection*{Node annotated recombination}

In the original (Griffiths) formulation, nodes which have multiple parents are
annotated with a breakpoint, representing a genomic position. The genome
lying to the left of that position is then taken to come from one parent, while
the genome lying to the right is taken to come from the other parent. This is
an elegantly concise encoding of the parental source, and is used, for example
in the ARGweaver .arg format. However, it suffers from three principal
drawbacks:

1. It assumes only a single breakpoint per node: problematic for representing
e.g. gene conversion or multiple chromosomes (but this could be circumvented by
having multiple breakpoints)
2. Nodes can only have a maximum of 2 parents. At
first sight, this seems a reasonable limitation, but it turns out that it can
be helpful to collapse multiple nodes into one, which potentially results in
more than two parents for any given node (see section XXXX)
3. Generating local
trees from such an encoding is time consuming (XXX describe *how* time
consuming - i,e. scaling properties with number of nodes - and why this is,
theoretically)

\subsection*{Edge annotated recombination}

An alternative approach to annotating nodes with the parent of origin of each
section of genome is to annotate each edge with the genomic region which is
inherited via that route. This approach was first introduced by Kelleher et al
(cite msprime paper) and now forms the basis of a comprehensive software
library.

% HYW: should discuss the link to biology & mitosis/meiosis: a "node" is
% strictly a cell - it is shorthand to associate it with an "individual"

@@

Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to 
model the evolution of viruses, reticulation nodes correspond to exchange of 
(non-recombining) genomic segments between viral particles inside a co-infected host. 
In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes 
correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations). 
The ARG is a subtype of a recombination network under this definition (restricting to one crossover 
recombination per reticulation node).

Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology. 
In \emph{tree-child networks}, every internal node must have at least one child that is a tree node; 
\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a 
tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on 
the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node). 
A reticulation loop that does not share any node with another reticulation loop is termed a gall 
\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is 
called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network 
by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number 
of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks, 
and the tractability afforded by their relatively simple structure results in a number of attractive 
combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of 
an ARG, without applying any of these restrictions.

% AI: have not mentioned any un-rooted networks (split, median, haplotype, reticulograms, etc).

@@

The problem of reconstructing ARGs for samples of recombining sequences has been of significant interest
essentially since the ARG was first defined. Early methods focussed on finding the most 
\emph{parsimonious} ARGs, i.e.\ those minimising the number of 
posited recombination events in the history of the sample \citep{hein1990reconstructing, song2003parsimonious,
lyngso2005minimum}. However, reconstructing a most parsimonious ARG for a given data set
is NP-hard \citep{wang2001perfect}, and in order to allow for larger input datasets,
subsequent parsimony-based methods have resorted to various relaxations and heuristics \citep{hein1993heuristic, 
song2005efficient, wu2008association, camara2016inference, thao2019hybrid, ignatieva2021kwarg}.
The scalability of these methods varies substantially based on the heuristic in use
and the underlying data structure, with the most efficient tools limited to analysing hundreds of sequences.
A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing
the most parsimonious ARGs, but rather aim for computational efficiency
while seeking to achieve good accuracy of the inferred genealogies
\citep{minichiello2006mapping, parida2008estimating, wu2011new, mirzaei2017rent}; some of these methods \citep{kelleher2019inferring, speidel2019method, zhang2021biobank} can be practically
used on megabase scale data and tens to hundreds of thousands of samples under human-like
parameters.
% AI: just to note, these papers do not all agree in their definition of an ARG. 
% I guess this is one of the points of the present paper, but should we mention this fact?
% Should we also mention that most of these infer just the topology, but some also the times?

@@

\begin{itemize}
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\item tskit can represent any ARG.
\end{itemize}

>>>>

\section*{The ARG as a data structure}
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}. These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?]
state space of ARGs. The focus subsequently shifted to
the more tractable---but still
NP-hard~\citep{wang2001perfect}---problem of computing
the miniumum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a data structure became less clear at this point,
with \citet{minichiello2006mapping} explicitly arguing for
the data structure interpretation.

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination, but
it has significant drawbacks as the basis for a general purpose
storage format.

[Not sure how to make a narrative out of this, but here are the
basic problems:]
\begin{enumerate}
\item The encoding is hamstrung by the interpretation of
nodes as \emph{events} of a particular type. We can either
have common ancestor or recombination events. This is true
in the coalescent, but is not in a more general Wright-Fisher
model.
\item We can also only have one crossover per recombination event.
This is unrealistic because things like gene conversion are common,
and means we can't have back and forth recombination between
two parents.
\item The passage of ancestral material along edges requires
extrinsic rules which must be shared by all implementations.
These rules can be quite restrictive when generating, and
in practise some have found it simpler to add explicit
annotations to the edges.
\item Generating local trees is inefficient, essentially requiring
$O(n T)$ time (or more?)
\end{enumerate}

@@

This ambiguity can be dealt with in a number of ways. In the original
treatments, the direction in which ancestral material travels at
a breakpoint is handled by a simple rule, sites to the left of the
breakpoint go to the left parent and sites to right go to the
right parent. [Illustrate the problem with simulating an ARG
like this]. ~\cite{ignatieva2021kwarg} resolve this problem by
associating a label ``P'' (prefix) or ``S'' (suffix) with the
outbound edges of a recombination event. This is the same approach
as used in beagle.

\section*{Anchoring the ARG to the pedigree}

The limitations of the Griffiths ARG encoding directly stem from its
rooting in the coalescent with recombination stochastic process.
Indeed, these limitations arise because it is such a \emph{good}
mathematical encoding, providing sufficient information to
define the realised ancestry under the assumptions of the model,
and no more.

We describe here an alternative formulation in which the
concepts are based on the diploid pedigree, which
is free from the limitations discussed in the previous section.
\cite{mathieson2020ancestry} and others [CITE] have also
suggested this approach, although without explicitly defining
the roles of nodes and the details of how ancestry
flows along edges.

@@

[Here are the basic components of the model]

\begin{enumerate}
\item A node corresponds to one of a diploid individual's two genomes.
This is linked explicitly to the pedigree, and is fully general.
Nodes \emph{may} be labelled with certain flags describing their properties,
but it's not necessary.
\item An edge between two nodes defines a genetic inheritance relationship
between two nodes. Edges are annotated with information about the
passage of ancestral materal. There are different forms these annotations
can take. For example, we can follow the approach taken in the Griffiths
ARG and just annotate recombinant edges with the interval $[0, x)$ or $[x, L)$,
depending on side of the breakpoint the edge is on. In the most
general case, we annotate each edge with the full set of disjoint
intervals of ancestral material that flows along that edge.
\citep{mcgill2013graphml} also proposed full ancestral material being
associated with each edge.
\end{enumerate}

@@

@@

Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
model the evolution of viruses, reticulation nodes correspond to exchange of
(non-recombining) genomic segments between viral particles inside a co-infected host.
In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
The ARG is a subtype of a recombination network under this definition (restricting to one crossover
recombination per reticulation node).
% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
% conversion (seems implied above)?

Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
A reticulation loop that does not share any node with another reticulation loop is termed a gall
\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
and the tractability afforded by their relatively simple structure results in a number of attractive
combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
an ARG, without applying any of these restrictions.
% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
% Also: are we not defining ARGs as "binary networks"? Might be good to specify this here.

@@

The problem of reconstructing ARGs for samples of recombining sequences has been of significant interest
essentially since the ARG was first defined. Early methods focussed on finding the most
\emph{parsimonious} ARGs, i.e.\ those minimising the number of
posited recombination events in the history of the sample \citep{hein1990reconstructing, song2003parsimonious,
lyngso2005minimum}. However, reconstructing a most parsimonious ARG for a given data set
is NP-hard \citep{wang2001perfect}, and in order to allow for larger input datasets,
subsequent parsimony-based methods have resorted to various relaxations and heuristics \citep{hein1993heuristic,
song2005efficient, wu2008association, camara2016inference, thao2019hybrid, ignatieva2021kwarg}.
The scalability of these methods varies substantially based on the heuristic in use
and the underlying data structure, with the most efficient tools limited to analysing hundreds of sequences.
A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing
the most parsimonious ARGs, but rather aim for computational efficiency
while seeking to achieve good accuracy of the inferred genealogies
\citep{minichiello2006mapping, parida2008estimating, wu2011new, mirzaei2017rent}; some of these methods \citep{kelleher2019inferring, speidel2019method, zhang2021biobank} can be practically
used on megabase scale data and tens to hundreds of thousands of samples under human-like
parameters.
% AI: just to note, these papers do not all agree in their definition of an ARG.
% I guess this is one of the points of the present paper, but should we mention this fact?
% Should we also mention that most of these infer just the topology, but some also the times?
% How much more description of the parsimony methods would be good? Maybe a mention of them splitting into
% "back in time" and "across local trees" types?

@@

\begin{itemize}
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\item tskit can represent any ARG.
\item Interchange is a real problem, and if we're going to realise the
potential for ARGs in genomics we must solve it. Efforts to standardise
phylogenetic networks haven't done very well (~\citep{cardona2008extended}
has 89 citations). The only real effort to standardise an ARG format
based on extending GraphML~\citep{mcgill2013graphml} has 11 citations.
GraphML~\citep{mcgill2013graphml} has some of the same key ideas
as tskit though, in particular allowing the annotation of ``live sites''
to an edge as a set of disjoint intervals that hold ancestral material.

\item We should discuss other types of ancestry network also.
The crucial distiction between an ARG and a phylogenetic network (I think?) is
the presence of position information specifying the location of
recombination events (and by extension, the complete passage
of ancestral material the graph). Without position information
we know that a particular node was an ancestor somewhere along the
sequence, but not exactly where.

\end{itemize}
